A Theory for Distributed Database Transactions

### 2019-06-14 2018_Transactions/articleTransactions.tex
Redone the storyline. Defined "action" and "system".

\documentclass{elsarticle}
\usepackage{graphicx}
%\usepackage{multicol}
%\usepackage{footmisc}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage[english]{babel}
%\usepackage[official,right]{eurosym}
\selectlanguage{english}
\include{preamramics}
\hyphenation{ExecEngine}
\newtheorem{lemma}{Lemma}
\def\id#1{\text{\em #1\/}}
\def\Events{{\mathit E}}
\begin{document}

@@

	We represent the ``history of a DIS up to some point in time'' by a relation $\id{prec}$ (pronounce: {\em precedes}),
	being a set of pairs from $\mathbb E$.
\[\id{prec}\subseteq\mathbb E\times\mathbb E\]
	Let $e_1\ \id{prec}\ e_2$ represent the fact that event $e_1$ has happened before event $e_2$,
	making $\id{prec}$ a relation on events in $\mathbb E$.
	We require:
\begin{eqnarray}
	&&(e,e)\notin\id{prec}\label{req:prec irreflexive}\\
	&&(e_1,e_2)\in\id{prec}^+\ \wedge\ (e_2,e_1)\in\id{prec}^+\ \Rightarrow\ e_1=e_2\label{req:precPlus asy}
\end{eqnarray}
	The events in the DIS, of which $\id{prec}$ represents the history, are 
\begin{eqnarray}
	&&\id{events}(\id{prec})\ =\ \{e_1|(e_1,e_2)\in\id{prec}\}\cup\{e_2|(e_1,e_2)\in\id{prec}\}\label{req:events}
\end{eqnarray}
	Requirement~\ref{req:prec irreflexive} excludes the situation that $e$ has happened before itself.
	To represent time faithfully, equation~\ref{req:precPlus asy} requires that the transitive closure $\id{prec}^+$ is antisymmetric.
	This makes $\id{prec}^*$ a reflexive, transitive, and antisymmetric relation, which lets us use it as a temporal order between events.
	So we shall call $\id{prec}$ "a history" because it represents all events that have happened up to a certain point in time and the (partial) order between them.
	Let $\mathbb H$ be the set of all conceivable histories, meaning that every $\id{prec}\in\mathbb H$.

@@

	The special case in which $\id{prec}^*$ is a total order is called a {\em sequential history}.
	Note that any nonempty finite sequential history has one current event and one initial event.
	Other finite histories may have more.
	In a sequential history structure, every event except the initial one has precisely one preceding event.
	In general, however, events may have multiple preceding events and multiple succeeding events.

	We define $\leq$ to represent the fact that history $\id{prec}$ lies in the past of history $\id{prec}'$.
\begin{definition}[$\leq$]
\label{def:lies in the past}
\begin{displaymath}
	\id{prec}\leq\id{prec}'\ \Leftrightarrow\ \id{prec}^*\subseteq\id{prec}'^*\ \wedge\ \id{initial}(\id{prec})\subseteq\id{initial}(\id{prec}')
\end{displaymath}
\end{definition}

\section{Traces}
\label{sct:Traces}
	A transaction in a DIS may be perceived as a distributed action,
	which leaves a trace of events in the history of a DIS.
	Let $\mathbb T$ be the set of all conceivable traces.
	With \id{prec} being a history, we define a trace $(e_1,e_2)_{\id{prec}}$ as follows:
\begin{definition}[Traces]
\label{def:Traces}
\item   $(e_1,e_2)_{\id{prec}}\in\mathbb T\ \Leftrightarrow\ (e_1,e_2)\in\text{prec}^*$
\end{definition}
	We use variables $x$, $y$, and $z$ to represent traces.
	We 
\begin{definition}[Empty trace]
\label{def:Empty trace}
\item   $\forall e: 0_{\id{prec}}\ =\ (e,e)_{\id{prec}}$
\end{definition}
	We denote the empty trace as $0$ and define $(e_1,e_1)=0=(e_2,e_2)$.
	If $(e_1,e)_{\id{prec}}$ and $(e,e_2)_{\id{prec}'}$ are traces then
	we define the concatenation operator $+:\mathbb T\rightarrow\mathbb T\rightarrow\mathbb T$ by
\begin{definition}[Concatenation of traces]
\label{def:concatenation}
\item   $(e_1,e)_{\id{prec}}+(e,e_2)_{\id{prec}'}=(e_1,e_2)_{\id{prec}\cup\id{prec}'}$
\end{definition}
	Note that the definition of $+$ is partial, because $(e_1,e_2)_{\id{prec}}+(e_3,e_4)_{\id{prec}'}$ is not defined unless $e_2=e_3$.
	Also, the result is defined in the combined history $\id{prec}^*\cup\id{prec}'^*$
	because $(e_1,e_2)\in(\id{prec}\cup\id{prec}')^*$. (proof needed)

	This definition of trace is an interpretation of trace algebra~\cite{Foster17b}, which describes discrete-event systems.
	Other interpretations of this algebra comprise real-time systems and hybrid systems, but that is beyond the scope of this paper.

	The concatenation operator $+$ and empty trace $0$ satisfy the axioms of trace algebras: (proof needed)
\begin{eqnarray}
	x+(y+z)&=&(x+y)+z\label{axi:associative}\\
	0+x&=&x+0\label{axi:lr-identity}\\
	x+y\ =\ x+z&\Rightarrow&y=z\label{axi:left cancellation}\\
	x+z\ =\ y+z&\Rightarrow&x=y\label{axi:right cancellation}\\
	x+z\ =\ 0&\Rightarrow&x=0\label{axi:no inverse}
\end{eqnarray}
	In terms of universal algebra, equations~\ref{axi:associative} and~\ref{axi:lr-identity} state that a trace algebra is a {\em monoid}.
	Equations~\ref{axi:left cancellation} and~\ref{axi:right cancellation} state that $+$ is {\em cancellative}.
	Equation~\ref{axi:no inverse} makes the cancellative monoid into a trace algebra.
	It implies that there is no inverse.

	The operator $\leq$ prefix, so we can define notions like shared history.
$
(e_1,e_2)\leq(e_3,e_4)
\Leftrightarrow
(\exists\id{prec}:(e_1,e_2)\in\id{prec}^*)\ \wedge\ (\exists\id{prec}':(e_3,e_4)\in\id{prec}'^*)\ \wedge\ \id{prec}\leq\id{prec}'
$
\begin{eqnarray}
	x\leq y&\Rightarrow&\exists z:\ x+z=y\label{def:Prefix}
\end{eqnarray}
	We also define subtraction of traces as a way to "cut off" a prefix from a trace.
\begin{definition}[Subtract]
\label{def:Subtract}
\begin{array}[t]{rcll}
	y-x&=&\text{if }x\leq y:&\exists z: y=x+z\Rightarrow z\\
	   & &\text{otherwise}:&0
\end{array}
\end{definition}

	Durability is meant to express that an action preserves history:
\begin{definition}[Durability]
\item An action $\id{act}:\mathbb H\rightarrow\mathbb H$ is durable iff for every $h\in\mathbb H$:
\label{def:Durability}
\begin{eqnarray}
	h\subseteq\id{act}(h)
\end{eqnarray}
\end{definition}

	An action $\id{act}_\id{prec}(e_1,e_2)$ is performed in isolation if it does not affect any other action in the system.
	This means that the only event in trace $(e_1,e_2)_\id{prec}$ that precedes events outside that trace is e_1.
	Likewise, the only event in trace $(e_1,e_2)_\id{prec}$ that is preceded by events outside that trace is e_2.
	To define this, we need an auxiliary function
	$\id{atomize}:\mathbb E\times\mathbb E\times\mathbb H\rightarrow\mathbb H$,
	which replaces a trace $(e_1,e_2)_\id{prec}$ by the pair $(e_1,e_2)$
	in a history $\id{prec}$.
\begin{eqnarray}
	\begin{array}[t]{r@{}l@{}l}
	\multicolumn{3}{l}{\id{atomize}(e_1,e_2,\id{prec})}\\
	   \ =\ \id{prec}-&(&\{ (e_1,e)\in\id{prec}| e_1\ \id{prec}^*\ e\ \wedge\ e\ \id{prec}^*\ e_2\}\ \cup\\
			  & &\{ (e,e_2)\in\id{prec}| e_1\ \id{prec}^*\ e\ \wedge\ e\ \id{prec}^*\ e_2\}\\
			  &)\\
		\cup&&\{ (e_1,e_2) \}\\
	\end{array}
\end{eqnarray}
	We can now define isolation as the property that the traces of the "atomized" history
	are the same as the atomization of all traces in that history.
\begin{definition}[Isolation]
\item An action $\id{act}_{\id{prec}}(e_1,e_2)$ works in isolation iff:
\label{def:Isolation}
\begin{eqnarray}
	(\id{atomize}(e_1,e_2,\id{prec}))^*&=&\id{atomize}(e_1,e_2,\id{prec}^*)
\end{eqnarray}
\end{definition}

@@

\begin{definition}[Atomic]
\item An action $\id{act}_{\id{prec}}(e_1,e_2)$ is atomic iff
\label{def:Atomicity}
\end{definition}

@@

\section{Transactions}
\label{sct:Transactions}
	In practice, a new state is computed from the current state by specifying a change, $\Delta$.
	Such changes are called {\em transactions}.
	To describe a DIS in terms of changes, we need operators to compute a new state ($s_2$) from a given state ($s_1$).
	In fact, we need two operators, $\oplus$ and $\ominus$, such that
\begin{equation}
	s_2\ =\ s_1\oplus(s_2\ominus s_1)
\label{req:oplus and ominus}
\end{equation}
	Let us call $s_2\ominus s_1$ the difference ($\Delta$) between states $s_2$ and $s_1$.
	We use $s_1\oplus\Delta$ to compute a new state from an existing state $s_1$ and a change $\Delta$.

>>>>

\documentclass{elsarticle}
\usepackage{graphicx}
%\usepackage{multicol}
%\usepackage{footmisc}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage[english]{babel}
%\usepackage[official,right]{eurosym}
\selectlanguage{english}
\include{preamramics}
\hyphenation{ExecEngine}
\newtheorem{lemma}{Lemma}
\def\id#1{\text{\it #1\/}}
\def\Events{{\mathit E}}
\begin{document}

@@

	We represent the ``history of a DIS up to some point in time'' by a relation $\id{prec}$ (pronounce: {\em precedes}),
	being a set of pairs from $\mathbb E$.
\[\id{prec}\subseteq\mathbb E\times\mathbb E\]
	Let $e_1\ \id{prec}\ e_2$ represent the fact that event $e_1$ has happened before event $e_2$,
	making $\id{prec}$ a relation on events in $\mathbb E$.
	We require:
\begin{eqnarray}
	&&(e,e)\notin\id{prec}\label{req:prec irreflexive}\\
	&&(e_1,e_2)\in\id{prec}^+\ \wedge\ (e_2,e_1)\in\id{prec}^+\ \Rightarrow\ e_1=e_2\label{req:precPlus asy}
\end{eqnarray}
	The events in the DIS, of which $\id{prec}$ represents the history, are 
\begin{eqnarray}
	&&\id{events}(\id{prec})\ =\ \{e_1|(e_1,e_2)\in\id{prec}\}\cup\{e_2|(e_1,e_2)\in\id{prec}\}\label{req:events}
\end{eqnarray}
	Requirement~\ref{req:prec irreflexive} excludes the situation that $e$ has happened before itself.
	To represent time faithfully, equation~\ref{req:precPlus asy} requires that the transitive closure $\id{prec}^+$ is antisymmetric.
	This makes $\id{prec}^*$ a reflexive, transitive, and antisymmetric relation, which lets us use it as a temporal order between events.
	We call $\id{prec}$ ``a history'' because it represents all events that have happened up to a certain point in time and the (partial) order between them.
	Let $\mathbb H$ be the set of all conceivable histories, meaning that every $\id{prec}\in\mathbb H$.

@@

	A history $\id{prec}$ where $\id{prec}^*$ is a total order is a special case, which we call a {\em sequential history}.
	Note that any nonempty finite sequential history has one current event and one initial event.
	Other finite histories may have more.
	In a sequential history structure, every event except the initial one has precisely one preceding event.
	In general, however, events may have multiple preceding events and multiple succeeding events.

	We define $\leq$ (called ``prefix'') to represent the fact that history $\id{prec}$ lies in the past of history $\id{prec}'$.
\begin{definition}[$\leq$]
\label{def:lies in the past}
\begin{displaymath}
	\id{prec}\leq\id{prec}'\ \Leftrightarrow\ \id{prec}^*\subseteq\id{prec}'^*\ \wedge\ \id{initial}(\id{prec})\subseteq\id{initial}(\id{prec}')
\end{displaymath}
\end{definition}

\section{Traces}
\label{sct:Traces}
	Transactions leave traces in the history of a DIS, so let us define traces.
	For this purpose we have chosen definitions that satisfy trace algebra
	and represent traces in discrete-event systems.
	Other interpretations of the same trace algebra comprise real-time systems and hybrid systems,
	which is proven by Foster~\cite{Foster17b}.
	We prove (TBD) that our definitions comply with Foster's axioms of trace algebra,
	but stick to one interpretation (discrete event systems).

	Let $\mathbb T$ be the set of all conceivable traces.
	With \id{prec} being a history, we define a trace $(e_1,e_2)_\id{prec}$ as follows:
\begin{definition}[Traces]
\label{def:Traces}
\item   $(e_1,e_2)_\id{prec}\in\mathbb T\ \Leftrightarrow\ (e_1,e_2)\in\id{prec}^*$
\end{definition}
	If it is obvious from the context which history ($\id{prec}$) is meant, we may write $(e_1,e_2)$ rather than $(e_1,e_2)_\id{prec}$.
	A trace algebra is a triple $\la\mathbb T,+,0\ra$,
	where the operator $+$ (concatenation) and empty trace $0$ satisfy the axioms of trace algebras:
\begin{eqnarray}
	x+(y+z)&=&(x+y)+z\label{axi:associative}\\
	0+x&=&x+0\label{axi:lr-identity}\\
	x+y\ =\ x+z&\Rightarrow&y=z\label{axi:left cancellation}\\
	x+z\ =\ y+z&\Rightarrow&x=y\label{axi:right cancellation}\\
	x+z\ =\ 0&\Rightarrow&x=0\label{axi:no inverse}
\end{eqnarray}
	In terms of universal algebra,
	equations~\ref{axi:associative} and~\ref{axi:lr-identity} state that a trace algebra is a {\em monoid}.
	Note that $+$ is {\em not} commutative.
	Equations~\ref{axi:left cancellation} and~\ref{axi:right cancellation} state that $+$ is {\em cancellative}.
	Equation~\ref{axi:no inverse} implies that there is no inverse.
	It makes the cancellative monoid into a trace algebra.

	For showing that definition~\ref{def:Traces} satisfies the axioms of trace algebra,
	we define the empty trace ($0$) and the trace concatenation ($+$):
\begin{definition}[Empty trace]
\label{def:Empty trace}
\item   $\forall e: 0_\id{prec}\ =\ (e,e)_\id{prec}$
\end{definition}
	We denote the empty trace as $0$ and define $(e_1,e_1)=0=(e_2,e_2)$.

	If $(e_1,e)_\id{prec}$ and $(e,e_2)_{\id{prec}'}$ are traces then
	we define the concatenation operator $+:\mathbb T\rightarrow\mathbb T\rightarrow\mathbb T$ by
\begin{definition}[Concatenation of traces]
\label{def:concatenation}
\item   $(e_1,e)_\id{prec}+(e,e_2)_{\id{prec}'}=(e_1,e_2)_{\id{prec}\cup\id{prec}'}$
\end{definition}
	Note that the definition of $+$ is partial, because $(e_1,e_2)_\id{prec}+(e_3,e_4)_{\id{prec}'}$ is not defined unless $e_2=e_3$.
	Nevertheless, the $+$ operator is closed: the result of composing two traces is always a trace.
	(proof needed)

	Let us define prefix for traces ($\leq$), to correspond closely to the prefix for histories
	(as given in definition~\ref{def:lies in the past}):
\begin{definition}[Prefix]
\label{def:Prefix of traces}
\item $\begin{array}[t]{rcl}
	(e,e_1)_\id{prec}\leq(e,e_2)_{\id{prec}'}&\Leftrightarrow&\id{prec}\leq\id{prec}'\ \wedge\ e_1\ \id{prec}'\ e_2
\end{array}$
\end{definition}
	We have defined this operator such that it satisfies Foster's algebraic definition of prefix~\cite{Foster17b}.
	(proof needed)
\begin{eqnarray}
	x\leq y&\Rightarrow&\exists z:\ x+z=y\label{def:Prefix}
\end{eqnarray}
	The prefix allows us to define subtraction of traces, to have a way to ``cut off'' an ``older'' part of a trace.
\begin{definition}[Subtract]
\label{def:Subtract}
\item $\begin{array}[t]{rcll}
	(e,e_1)_\id{prec}-(e,e_2)_{\id{prec}'}&=&(e_1,e_2)_{\id{prec}'}&\text{if }(e,e_1)_\id{prec}\leq(e,e_2)_{\id{prec}'}\\
	   &=&0&\text{otherwise}
\end{array}$
\end{definition}
	This definition satisfies Foster's definition of subtract (eqn.~\ref{axi:Foster subtract})
\begin{equation}
\label{axi:Foster subtract}
\begin{array}[t]{rcll}
	y-x&=&z&\text{if }x\leq y\ \wedge\ \exists z: y=x+z\\
	   &=&0&\text{otherwise}
\end{array}
\end{equation}

\section{State}
\label{sct:State}
	Typically, distributed information systems are stateful.
	Every event in the history of a DIS may cause a different (local) state.
	(The distributed nature of a DIS means that states are local.)
	Let $\mathbb S$ be the set of all conceivable states.
	So we assume a function $\id{state}_\id{prec}:\mathbb E\rightarrow\mathbb S$ that assigns one state to every event $e$ in history $\id{prec}$.
	State $\id{state}_\id{prec}(e)$ represents the state immediately after $e$ has happened.
	We assume an empty state, $\emptyset$, which is the state immediately before an initial event has happened.
	The global state of the system, if relevant, can be computed by $\{\id{state}_\id{prec}(e) | e\in\id{current}(\id{prec})\}$.
	Locally we work with one specific current event, $e$, with $\id{state}_\id{prec}(e)$ representing the local current state.

	In a DIS, new states typically emerge as a change on an existing state.
	So we want to describe a DIS in terms of changes.
	Let $\mathbb D$ be the set of all conceivable changes and let $\delta_\id{prec}:\mathbb E\times\mathbb E\rightarrow\mathbb D$
	be a function that maps a pair of events to the corresponding change.
	The change between the states of events $e_1$ and $e_2$ is given by $\delta_\id{prec}(e_1,e_2)$.
	Let us define a DIS as a tuple $(\id{prec},\delta_\id{prec})$,
	in which $\id{prec}$ is a history and
	$\delta_\id{prec}$ represents the changes of state that have taken place throughout the history.

	We need operators to compute a new state ($s_2$) from a given state ($s_1$).
	In fact, we need two operators, $\oplus:\mathbb S\times\mathbb D\rightarrow\mathbb S$ and $\ominus:\mathbb S\times\mathbb S\rightarrow\mathbb D$, such that
\begin{equation}
	s_2\ =\ s_1\oplus(s_2\ominus s_1)
\label{req:oplus and ominus}
\end{equation}
	Let us call $s_2\ominus s_1$ the difference between states $s_2$ and $s_1$.
	The following equations represent the laws about state.
	They must be kept true all the time in a DIS.
\begin{eqnarray}
	\delta_\id{prec}(e_1,e_2)&=&\id{state}_\id{prec}(e_2)\ominus\id{state}_\id{prec}(e_1)\\
	e_1\ \id{prec}\ e_2&\Rightarrow&\id{state}_\id{prec}(e_2)\ =\ \id{state}_\id{prec}(e_1)\oplus\delta_\id{prec}(e_1,e_2)
\end{eqnarray}

\section{Transactions}
\label{sct:Transactions}
	With the theory in place, let us proceed to define transactions.
	Let us first define the notion of action ($\id{act}$)
	and then define the classical properties of transactions: atomic, consistent, isolated, and durable.

	An action is denoted as $\id{act}_{e_1,e_2,\Delta}$, which is a mapping from one DIS to another
\begin{equation}
\begin{array}{rl}
	\multicolumn{2}{l}{e_2\not\in\id{events}(\id{prec})\ \wedge\ (e_1,e_2)\in\id{prec}'^*\ \Rightarrow}\\
	&\id{act}_{e_1,e_2,\Delta}(\id{prec},\delta_\id{prec})\ =\ (\id{prec}',\delta'_{\id{prec}'})\\
	&\text{where}\\
	&\ \ \begin{array}[t]{rcll}
		\delta'_{\id{prec}'}(e,e')&=&\Delta&\text{if }(e,e')=(e_1,e_2)\\
		\delta'_{\id{prec}'}(e,e')&=&\delta_\id{prec}(e,e')&\text{otherwise}\\
		\end{array}
\end{array}
\end{equation}
	The state function can be computed recursively as follows:
\begin{equation}
\begin{array}{rcll}
	\id{state}_\id{prec}(e)&=&\id{state}_\id{prec}(e_1)\oplus\delta(e_1,e)&\text{if }(e_1,e)\in\id{prec}\\
	                       &=&\emptyset&\text{otherwise}
\end{array}
\end{equation}

	Durability is meant to express that an action preserves history.
\begin{definition}[Durability]
\item An action $\id{act}:\mathbb H\rightarrow\mathbb H$ is durable iff for every $h\in\mathbb H$:
\label{def:Durability}
\begin{eqnarray}
	\id{prec}^*\subseteq\id{prec}'^*
\end{eqnarray}
\end{definition}

	An action $\id{act}_\id{prec}(e_1,e_2)$ is performed in isolation if it does not affect any other action in the system.
	This means that the only event in trace $(e_1,e_2)_\id{prec}$ that precedes events outside that trace is $e_1$.
	Likewise, the only event in trace $(e_1,e_2)_\id{prec}$ that is preceded by events outside that trace is $e_2$.
	To define this, we need an auxiliary function
	$\id{atomize}:\mathbb E\times\mathbb E\times\mathbb H\rightarrow\mathbb H$,
	which replaces a trace $(e_1,e_2)_\id{prec}$ by the pair $(e_1,e_2)$
	in a history $\id{prec}$.
\begin{eqnarray}
	\begin{array}[t]{r@{}l@{}l}
	\multicolumn{3}{l}{\id{atomize}(e_1,e_2,\id{prec})}\\
	   \ =\ \id{prec}-&(&\{ (e_1,e)\in\id{prec}| e_1\ \id{prec}^*\ e\ \wedge\ e\ \id{prec}^*\ e_2\}\ \cup\\
			  & &\{ (e,e_2)\in\id{prec}| e_1\ \id{prec}^*\ e\ \wedge\ e\ \id{prec}^*\ e_2\}\\
			  &)\\
		\cup&&\{ (e_1,e_2) \}\\
	\end{array}
\end{eqnarray}
	We can now define isolation as the property that the traces of the "atomized" history
	are the same as the atomization of all traces in that history.
\begin{definition}[Isolation]
\item An action $\id{act}_\id{prec}(e_1,e_2)$ works in isolation iff:
\label{def:Isolation}
\begin{eqnarray}
	(\id{atomize}(e_1,e_2,\id{prec}))^*&=&\id{atomize}(e_1,e_2,\id{prec}^*)
\end{eqnarray}
\end{definition}

@@

\begin{definition}[Atomic]
\item An action $\id{act}_\id{prec}(e_1,e_2)$ is atomic iff
\label{def:Atomicity}
\end{definition}

@@

\section{Transactions}
\label{sct:Transactions}
	In practice, a new state is computed from the current state by specifying a change, $\Delta$.
	Such changes are called {\em transactions}.
	We use $s_1\oplus\Delta$ to compute a new state from an existing state $s_1$ and a change $\Delta$.
